Generating Random Logic Programs Using Constraint Programming

### 2020-01-21 paper/paper.tex
Another plot

\usetikzlibrary{arrows.meta}
\Crefname{constraint}{Constraint}{Constraints}
\relpenalty=10000
\binoppenalty=10000
\def\multiset#1#2{\ensuremath{\left(\kern-.3em\left(\genfrac{}{}{0pt}{}{#1}{#2}\right)\kern-.3em\right)}}

@@

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (or) at (-0.375, 0) {$\lor$};
    \node[draw,circle,gray,text=black] (not) at (-1.5, -1) {$\neg$};
    \node[draw,circle,gray,text=black] (and) at (0.75, -1) {$\land$};
    \node[draw,gray,text=black] (P) at (-1.5, -2) {$\mathsf{P}(X)$};
    \node[draw,gray,text=black] (Q) at (0, -2) {$\mathsf{Q}(X)$};
    \node[draw,gray,text=black] (R) at (1.5, -2) {$\mathsf{P}(X)$};
    \draw[gray] (or) -- (not);
    \draw[gray] (or) -- (and);
    \draw[gray] (not) -- (P);
    \draw[gray] (and) -- (Q);
    \draw[gray] (and) -- (R);
  \end{tikzpicture}
  \caption{A tree representation of the formula from \cref{example:formula}}
  \label{fig:example_tree}
\end{figure}

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (and) at (0, 0) {$\land$};
    \node[draw,circle,gray,text=black] (not) at (1.5, -1) {$\neg$};
    \node[draw,gray,text=black] (parent) at (-1.5, -1) {$\mathsf{parent}(X, Y)$};
    \node[draw,gray,text=black] (mother) at (1.5, -2) {$\mathsf{mother}(X, Y)$};
    \draw[gray] (and) -- (parent);
    \draw[gray] (and) -- (not);
    \draw[gray] (not) -- (mother);
  \end{tikzpicture}
  \caption{The tree representation of the body of Clause~\eqref{eq:example_clause}.}
  \label{fig:example_tree2}
\end{figure}
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
    \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
    \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
    \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
    \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
    \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
    \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
  \end{tikzpicture}
  \caption{The predicate dependency graph of the program in \cref{ex:program}.
    Positive edges are labeled with `$+$', and negative edges with `$-$'.}
  \label{fig:predicate_dependencies}
\end{figure}

\begin{example} \label{ex:program}
  Consider the following (fragment of a) program:
  \begin{align}
    \mathsf{sibling}(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z), \nonumber \\
    \mathsf{father}(X, Y) &\gets \mathsf{parent}(X, Y) \land \neg\mathsf{mother}(X, Y). \label{eq:example_clause}
  \end{align}
  Its predicate dependency graph is in \cref{fig:predicate_dependencies}.
  Because of the negation in \cref{eq:example_clause} (as seen in
  \cref{fig:example_tree2}), the edge from $\mathsf{mother}$ to
  $\mathsf{father}$ is negative, while the other two edges are positive.

@@

  \caption{Propagation for independence}

@@

  \caption{Entailment for independence}

@@

  \caption{Dependencies of a predicate}

@@

  \caption{The adjacency matrix defined using \cref{def:adjacency_matrix} for

@@

@@

  \caption{The predicate dependency graph that corresponds to
    \cref{fig:dependencies_matrix}. Dashed edges are undetermined---they may
    or may not exist.}

@@

\begin{algorithm}
  \KwData{a program $\mathscr{P}$}
  \SetKwFunction{hasNegativeCycles}{hasNegativeCycles}
  Let $\mathscr{R} \subseteq \mathscr{P}$ be the largest subprogram of
  $\mathscr{P}$ with its $\variable{structure}$ and predicates in both body
  and head fully determined\footnotemark\;
  \lIf{$\mathscr{R} = \emptyset$}{\Return{\textsc{undefined}}}
  \If{\hasNegativeCycles($G_{\mathscr{R}}$)}{
    \Return{\textsc{false}}\;
  }
  \lIf{$\mathscr{R} = \mathscr{P}$}{\Return{\textsc{true}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment check for negative cycles}
  \label{alg:negative_cycles}
\end{algorithm}
\footnotetext{The arguments (whether variables or constants) are irrelevant to
  our definition of independence.}

@@

\begin{figure*}[t]
  \centering
  \input{phase_transition.tex}
  \caption{TODO}
\end{figure*}

@@

% TODO: consider replacing \maxNumClauses{} with max_num_additional_clauses
\begin{itemize}
\item $|\predicates{}|$, $|\variables{}|$, $|\constants{}|$, $\maxNumNodes{}$,
  the number of additional clauses are in $\{ 1, 2, 4, 8 \}$.
\item $\maxArity{} \in \{ 1, 2, 3, 4 \}$.
\item all possible numbers of independent pairs: $0, \dots,
  \binom{|\predicates{}|}{2}$.
\end{itemize}
Repeat 10 times:
\begin{itemize}
\item Pick a random combination of $|\predicates{}|$ arities such that
  $\maxArity{}$ is occurs at least once (in an ordered way, since order doesn't
  matter).
\item Pick the required number of independent pairs (without repetition).
\item Run the solver with a \SI{60}{\second} timeout.
\end{itemize}

Results:
\begin{itemize}
\item almost \num{400000} experiments.
\item \SI{97.7}{\percent} of total running times is less than \SI{1}{\second}
\item 4 timeouts: all with 8 predicates, 8 additional clauses, and
  $\maxNumNodes{} = 8$; the rest of the parameters differ.
\item Running time (or number of nodes) as a function of various parameters:
  \begin{itemize}
  \item The number of clauses has a clear impact on time but not number of
    nodes.
  \item $\maxNumNodes{}$ has a huge impact on both.
  \item The proportion of independent pairs of predicates is acting weird.
    Investigate.
  \end{itemize}
\end{itemize}

>>>>

\Crefname{constraint}{Constraint}{Constraints}
\Crefname{clause}{Clause}{Clauses}
\creflabelformat{clause}{#2(#1)#3}

\usetikzlibrary{arrows.meta}
\relpenalty=10000
\binoppenalty=10000
\def\multiset#1#2{\ensuremath{\left(\kern-.3em\left(\genfrac{}{}{0pt}{}{#1}{#2}\right)\kern-.3em\right)}}

@@

% UAI format: section/subsection titles are in all caps, figure/table titles are
% in TitleCase, table caption above the table

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (or) at (-0.375, 0) {$\lor$};
    \node[draw,circle,gray,text=black] (not) at (-1.5, -1) {$\neg$};
    \node[draw,circle,gray,text=black] (and) at (0.75, -1) {$\land$};
    \node[draw,gray,text=black] (P) at (-1.5, -2) {$\mathsf{P}(X)$};
    \node[draw,gray,text=black] (Q) at (0, -2) {$\mathsf{Q}(X)$};
    \node[draw,gray,text=black] (R) at (1.5, -2) {$\mathsf{P}(X)$};
    \draw[gray] (or) -- (not);
    \draw[gray] (or) -- (and);
    \draw[gray] (not) -- (P);
    \draw[gray] (and) -- (Q);
    \draw[gray] (and) -- (R);
  \end{tikzpicture}
  \caption{A Tree Representation of the Formula From \cref{example:formula}}
  \label{fig:example_tree}
\end{figure}

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (and) at (0, 0) {$\land$};
    \node[draw,circle,gray,text=black] (not) at (1.5, -1) {$\neg$};
    \node[draw,gray,text=black] (parent) at (-1.5, -1) {$\mathsf{parent}(X, Y)$};
    \node[draw,gray,text=black] (mother) at (1.5, -2) {$\mathsf{mother}(X, Y)$};
    \draw[gray] (and) -- (parent);
    \draw[gray] (and) -- (not);
    \draw[gray] (not) -- (mother);
  \end{tikzpicture}
  \caption{A Tree Representation of the Body of \cref{eq:example_clause}}
  \label{fig:example_tree2}
\end{figure}
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
    \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
    \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
    \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
    \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
    \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
    \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
  \end{tikzpicture}
  \caption{The Predicate Dependency Graph of the Program In \cref{ex:program}.
    Positive Edges Are Labelled with `$+$', and Negative Edges with `$-$'.}
  \label{fig:predicate_dependencies}
\end{figure}

\begin{example} \label{ex:program}
  Consider the following (fragment of a) program:
  \begin{align}
    \mathsf{sibling}(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z), \nonumber \\
    \mathsf{father}(X, Y) &\gets \mathsf{parent}(X, Y) \land \neg\mathsf{mother}(X, Y). \label[clause]{eq:example_clause}
  \end{align}
  Its predicate dependency graph is in \cref{fig:predicate_dependencies}.
  Because of the negation in \cref{eq:example_clause} (as seen in
  \cref{fig:example_tree2}), the edge from $\mathsf{mother}$ to
  $\mathsf{father}$ is negative, while the other two edges are positive.

@@

  \caption{Propagation for Independence}

@@

  \caption{Entailment for Independence}

@@

  \caption{Dependencies of a Predicate}

@@

  \caption{The Adjacency Matrix Defined Using \cref{def:adjacency_matrix} For

@@

\begin{figure}[t]

@@

  \caption{The Predicate Dependency Graph That Corresponds To
    \cref{fig:dependencies_matrix}. Dashed Edges Are Undetermined---They May
    Or May Not Exist.}

@@

\begin{algorithm}
  \KwData{a program $\mathscr{P}$}
  \SetKwFunction{hasNegativeCycles}{hasNegativeCycles}
  Let $\mathscr{R} \subseteq \mathscr{P}$ be the largest subprogram of
  $\mathscr{P}$ with its $\variable{structure}$ and predicates in both body
  and head fully determined\footnotemark\;
  \lIf{$\mathscr{R} = \emptyset$}{\Return{\textsc{undefined}}}
  \If{\hasNegativeCycles($G_{\mathscr{R}}$)}{
    \Return{\textsc{false}}\;
  }
  \lIf{$\mathscr{R} = \mathscr{P}$}{\Return{\textsc{true}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment Check for Negative Cycles}
  \label{alg:negative_cycles}
\end{algorithm}
\footnotetext{The arguments (whether variables or constants) are irrelevant to
  our definition of independence.}

@@

@@

% TODO: consider replacing \maxNumClauses{} with max_num_additional_clauses
\begin{itemize}
\item $|\predicates{}|$, $|\variables{}|$, $|\constants{}|$, $\maxNumNodes{}$,
  $\maxNumClauses{} - |\predicates{}|$ are in $\{ 1, 2, 4, 8 \}$.
\item $\maxArity{} \in \{ 1, 2, 3, 4 \}$.
\item all possible numbers of independent pairs: $0, \dots,
  \binom{|\predicates{}|}{2}$.
\end{itemize}
Repeat 10 times:
\begin{itemize}
\item Pick a random combination of $|\predicates{}|$ arities such that
  $\maxArity{}$ is occurs at least once (in an ordered way, since order doesn't
  matter).
\item Pick the required number of independent pairs (without repetition).
\item Run the solver with a \SI{60}{\second} timeout.
\end{itemize}

Results:
\begin{itemize}
\item almost \num{400000} experiments, almost \num{300000} of which are with
  $|\predicates{}| = 8$ (since the maximum number of independent pairs grows
  with $|\predicates{}|$).
\item \SI{97.7}{\percent} of total running times is less than \SI{1}{\second}
\item 4 timeouts: all with 8 predicates, 8 additional clauses, and
  $\maxNumNodes{} = 8$; the rest of the parameters differ.
\item Running time (or number of nodes) as a function of various parameters:
  \begin{itemize}
  \item The number of clauses has a clear impact on time but not number of
    nodes.
  \item $\maxNumNodes{}$ has a huge impact on both.
  \item The proportion of independent pairs of predicates is acting weird.
    Investigate.
  \end{itemize}
\end{itemize}

% TODO: describe the phase transition in this figure. still, almost 300k
% data points. A phase transition in mean but not
% median, i.e., some instances become really hard, but most remain easy.
\begin{figure*}[t]
  \centering
  \input{phase_transition.tex}
  \caption{The Distribution of the Number of Nodes in the Binary Search Tree as
    a Function of the Number of Independent Pairs of Predicates for
    $|\predicates| = 8$. Significant Outliers Are Hidden, The Dots Denote Mean
    Values, and the Vertical Axis Is on a $\log_{10}$ Scale.}
\end{figure*}

\begin{figure}[t]
  \centering
  \input{impact.tex}
  \caption{The Mean Number of Nodes in the Binary Search Tree for Each Value of
    Each Experimental Parameter. Note That the Horizontal Axis Is on a $\log_2$
    Scale.}
  \label{fig:impact}
\end{figure}

Observations about \cref{fig:impact}:
\begin{itemize}
\item The same plot for the median is very similar.
\item The growth of each curve suggest how well/poorly the model scales with
  higher values of the parameter.
\item It is clear that the number of nodes that can represent a body of a clause
  is the limiting factor. A big tree significantly increases the search space,
  and it becomes harder to ensure that there are no negative cycle and 
\end{itemize}
